Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-18 doc/invt.tex
separate invt_notes.tex

By assuming Eq. \eqref{eq:w_detailedbalance},
we can diagonalize $\mathbf w$ with a set of
eigenvectors
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 1}^n
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, p_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { p_i }
  &=
  \delta_{kl}
  .
\label{eq:eig_orthonormal_rows}
\end{align}
%
\note{We can define a symmetric matrix $\hat{\mathbf w}$
as $\hat w_{ij} = \sqrt{p_i/p_j} \, w_{ij}$,
which can be diagonalized\cite{vankampen}
with a set of orthonormal eigenvectors:
%
$\sum_{i = 1}^n \varphi_{ki} \, \hat w_{ij} = \lambda_k \, \varphi_{kj}$.
%
Then,
in terms of diagonalizing $\mathbf w$, we have
$$\phi_{ij} \equiv \sqrt{p_j} \, \varphi_{ij}$$
satisfying Eqs. \eqref{eq:eig_orthonormal_cols}
and \eqref{eq:eig_orthonormal_rows}.
%
}%

@@

In analogous to Eq. \eqref{eq:x2t_average},
we can write,
from Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:y2_sum},
the error after a period $T$,
%
\begin{equation}
  \Err
  =
  \Err_R + \Err_A,
  \label{eq:error_split}
\end{equation}
%
as the sum of the residual error
(from the decay of the initial error)
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
\begin{equation}
  \Err_A
  =
  \int_0^T \int_0^T
  \sum_{k = 2}^n
    \dot u_k\bigl( q(t) \bigr) \,
    \dot u_k\bigl( q(t') \bigr) \,
    \kappa_k(t - t') \, dt \, dt',
  \label{eq:error_asym}
\end{equation}
%
where
\begin{align*}
  u_k(q')
  &\equiv
  e^{\lambda_k \, [q' - q(T)]},
  \\
  \kappa_k(t - t')
  &\equiv
  \left\langle
    \theta_k(t) \, \theta_k(t')
  \right\rangle.
\end{align*}
%
\note{The extremal condition of $\Err_A$ is given by
%
\begin{align}
  \sum_{k = 2}^n
  \lambda_k \, u_k\bigl( q(t) \bigr)
  \int_0^T
    \ddot u_k \bigl( q(t') \bigr) \,
    \kappa_k(t' - t) \, dt' = 0.
  \tag{N1}
\label{eq:optimal_mbin}
\end{align}
%
If all $\lambda_k$, hence all $u_k$, are the same,
the above condition can be satisfied with $\ddot u_k = 0$,
which recovers the single-bin case.
%
Unfortunately, the above expression
appears to be useless in general.
}

@@

In the asymptotic regime
(i.e., if the simulation length, $T$, is sufficiently long),
we can approximate $\kappa_k(t)$ as
%
\begin{equation}
  \kappa_k(t) \approx \Gamma_k \, \delta(t)
  ,
\label{eq:kappa_delta}
\end{equation}
%
where
\begin{equation}
  \Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_k = \sum_{t = -\infty}^\infty \kappa_k(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_asym} is simplified as
%
\begin{align}
  \Err_A
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt.
  \label{eq:error_asym1}
\end{align}
%
\note{The next order correction to Eq. \eqref{eq:kappa_delta}
would lead to a correction to $\Err$
that is about $\alpha_{\max}^2$ times as large,
where $\alpha_{\max} = \max \alpha(t)$.
%
This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_k(\omega) = \Gamma_k
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_k(\omega) = \Gamma_k + \Gamma^{(2)}_k \omega^2 + \dots,
  $$
  which leads to a correction to $\Err$
  $$
  \Gamma^{(2)}_k
  \int_0^T \ddot u_k^2\bigl( q(t) \bigr) \, dt
  \propto
  \alpha^2 \, \Err.
  $$
  This correction is of order $\alpha^2$
  compared to the main contribution in Eq. \eqref{eq:error_asym1}.
  %
  Thus, we can assume Eq. \eqref{eq:kappa_delta},
  once $\alpha(t)$ has dropped below a certain level.
}%
\note{
  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{k = 1}^n
      \Gamma_k \, \lambda_k \,
      u_k\bigl( q(t) \bigr) \,
      \ddot u_k\bigl( q(t) \bigr) = 0
    .
    \tag{N2}
    \label{neq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_k$ as the position,
  and $\Gamma_k \, \lambda_k$ as the mass,
  then the right-hand side gives the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}%
In this way,
the underlying sampling process
affects the error only through the set of numbers, $\Gamma_i$'s.

@@

We shall use this result to model the initial error
before starting the schedule $\alpha(t)$ to be optimized.
%
That is, we assume
$\alpha(t) = \alpha_0$
for $t < 0$, and
the system has entered the equilibrium
under a fixed $\alpha_0$ at $t = 0$.
%
We can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.
%
\note{Another possible approximation
  is to interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k(t) = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}
%
Then,
the residual error at the end of period $T$
is given by
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
      \frac 1 2 \, \alpha_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
Note that the residual error
depends on $\alpha(t)$ only through
the integral $q(T) = \int_0^T \alpha(t) \, dt$.
%
Thus, if we variate $\alpha(t)$ under a fixed $q(T)$,
this residual part can be ignored,
and the optimization can be done exclusively
on the asymptotic part, $\Err_A$.

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
%\label{eq:error_asym_Lagrangian}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  ,
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
  %\label{eq:q_opt}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in th above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

Unfortunately,
Eq. \eqref{eq:q_opt} is not an explicit equation of $\alpha(t)$.
%
However,
by differentiating Eq. \eqref{eq:Lagrangian_const},
we have
%
\begin{equation}
  \frac{ d   }
       { d t }
  \left[
    \frac{       1     }
         { \alpha( t ) }
  \right]
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  }
  \equiv
  \lambda(t)
  .
  \label{eq:dinvadt}
\end{equation}
%
The expression in the middle is an average
of the eigenvalues, $\lambda_k$'s,
weighted by $\Gamma_k \, \lambda_k^2 \, u_k^2$.
%
We shall refer to the average as the instantaneous
eigenvalue, $\lambda(t)$,
which gives the rate of change of $\alpha^{-1}(t)$.

\note{For the single-bin scheme, $\lambda_k = 1$,
and $\lambda(t)$ is always unity.
%
So,
$$
\frac { d   }
      { d t }
\left[
  \frac{      1    }
       { \alpha(t) }
\right]
=
1,
$$
which again leads to Eq. \eqref{eq:alpha_invt1}.

For a general multiple-bin scheme,
we can characterize the solution of
Eq. \eqref{eq:dinvadt} as
\begin{equation}
  \alpha(t)
  =
  \frac{                  1                     }
       { \lambda(t) \, \bigl[ t + t_0(t) \bigr] }
  ,
\notag
%\label{eq:alpha_approxinvt}
\end{equation}
%
where
$t_0(t)$ is a slowly-varying function.
}

By further differentiating Eq. \eqref{eq:dinvadt},
we can show that $\lambda(t)$
is an increasing function.
%
\note{Proof that $\lambda(t)$ is an increasing function.
  Using $\dot u_k = \lambda_k \, u_k \, \alpha(t)$,
  we have
  $$
  \begin{aligned}
    \dot \lambda
    &=
    \frac
    {
      \sum_{k = 2}^n 2 \, \Gamma_k \, \lambda_k^3 \, u_k \, \dot u_k
    }
    {
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2
    }
    -
    \frac
    {
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^3 \, u_k^2
      \sum_{k = 2}^n 2 \, \Gamma_k \, \lambda_k^2 \, u_k \, \dot u_k
    }
    {
      \left(
        \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2
      \right)^2
    }
    \\
    &=
    \frac
    {
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^4 \, u_k^2
      \,
      \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2
      -
      \left(
        \sum_{k = 2}^n \Gamma_k \, \lambda_k^3 \, u_k^2
      \right)^2
    }
    {
      \left(
        \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2
      \right)^2
    }
    2 \alpha(t).
  \end{aligned}
  $$
  The numerator of the fraction of the last expression
  is nonnegative by the Cauchy-Schwarz inequality,
  or by the nonpositivity of the discriminant
  of the following nonnegative definite quadratic form of $X$:
  $$
  \sum_{k = 2}^n \Gamma_k \, \lambda_k^2 \, u_k^2 \, (X - \lambda_k)^2 \ge 0.
  $$
  Thus, $\dot \lambda \ge 0$, and $\lambda(t)$ is monotonically increasing.
}%
%
Below we shall study the initial and final values.
%
At the end of the simulation,
$t = T$, we have $u_k = 1$, and
%
\begin{equation}
  \lambda(T)
  =
  \frac{
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^3
  }
  {
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
  }
  ,
\notag
%\label{eq:dinvadt_limit2}
\end{equation}
is independent of the simulation length, $T$.
%
Further, $\Gamma_k$ often increases with $\lambda_k$,
such that the relative weight $\Gamma_k \, \lambda_k^2$
also increases with $\lambda_k$.
%
It follows that $\lambda(T)$ will be dominated
by the larger values of $\lambda_k$'s.

By contrast,
at the beginning of the simulation,
the relative weight, $\Gamma_k \, \lambda_k^2 \, u_k^2(0)$,
is heavily influenced by the exponential factor
$u_k^2(0) = e^{ -2 \, \lambda_k \, q(T) }$.
%
Thus, $\lambda(0)$
is sensitive to the simulation length, $T$.
%
A longer simulation means a smaller $\lambda(0)$,
and that $\lambda(t)$ would stay small
for longer near the beginning of the simulation.

@@

Parallel to Eq. \eqref{eq:phi_pbc},
we define the out-of-boundary values
of $\phi_i$ as
%
\begin{equation}
  \phi_i
  =
  \begin{dcases}
    \phi_{ 1 - i }           & \mathrm{for \;} i \le 0, \\
    \phi_{ 2 \, n + 1 - i }  & \mathrm{for \;} i > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that Eq. \eqref{eq:wmul_to_convol}
still holds.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}%
Thus, we also have $n$ orthonormal eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
compatible with Eq. \eqref{eq:phi_refl},
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, \pi}
            {                    n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
with eigenvalues
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \pi}{n}
  .
\label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{Derivation.
  First consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies the reflective boundary condition,
  Eq. \eqref{eq:phi_refl}.
  %
  So, by Eq. \eqref{eq:wmul_to_convol},
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue_refl}.

  \hrulefill

  Next, let us compute the normalization factor.
  %
  If $k = 1$, $\Phi^{(1)}_i = 1$, and
  $
  \sum_{i = 1}^n \left( \Phi^{(1)}_i \right)^2 = n.
  $
  For $k \ge 2$,
  $$
  \begin{aligned}
    \sum_{i = 1}^n \cos^2 \left[\left(i - \frac1 2 \right) a\right]
    &=
    \frac n 2
    +
    \frac 1 2
    \sum_{i = 1}^n \cos\left[(2 i - 1)\, a \right]
    =
    \frac n 2,
  \end{aligned}
  $$
  where $a = \frac{k-1}{n} \pi$.
  Thus, the normalization factor $\sqrt{(2 - \delta_{k1})/n}$
  encompasses both cases.

  \hrulefill

  To show the column-wise orthogonality,
  we need a lemma, for $a = q \, \pi/n$,
  with $q$ a positive integer.
  $$
  \begin{aligned}
  \sum_{k = 1}^n \cos[(k - 1) \, a]
  &=
  1 + \cos a + \dots + \cos[(n - 1) \, a]
  \\
  &=
  \frac{
        \sin\frac a 2
      + \sin \left[ \left( n - \frac 1 2 \right) a \right]
      }
      {
        2 \, \sin \frac a 2
      }
  \\
  &=
  \frac{ 1 - (-1)^q } { 2 }
  = \operatorname{odd}(q),
  \end{aligned}
  $$
  %
  where, we have used
  $\sin \left[ \left( n - \frac 1 2 \right) a \right]
  = \sin \left( q \, \pi - \frac a 2 \right)
  = -(-)^q\sin\frac a 2.$
  %
  and $\operatorname{odd}(q)$
  yields $1$ for odd $q$ or $0$ otherwise.
  %
  Adding the case of $q = 0$, where the sum yields $n$,
  we get
  $$
  \sum_{k = 1}^n
    \cos\left[(k - 1) \, \frac { q \, \pi } { n }  \right]
  = n \, \delta_{q, 0}
  + \operatorname{odd}(q).
  $$

  Now, for the orthogonality,
  $$
  \begin{aligned}
    \sum_{k = 1}^n
    \Phi^{(k)}_i \, \Phi^{(k)}_j
    &=
    \frac 1 2
    \sum_{k = 1}^n
    \left[
      \cos \tfrac{ (i - j) (k - 1) \, \pi }
                 {         n              }
      +
      \cos \tfrac{ (i + j - 1) (k - 1) \, \pi }
                 {             n              }
    \right]
    \\
    &=
    \frac 1 2
    \left[
      n \, \delta_{i, j}
      +
      \operatorname{odd}(i - j)
      +
      \operatorname{odd}(i + j - 1)
    \right]
    \\
    &=
    \frac n 2 \, \delta_{i, j}
    + \frac 1 2.
  \end{aligned}
  $$
  where we have used the fact
  that $i - j$ shares the parity with $i + j$
  in the last step.
  %
  So
  $$
    \sum_{k = 1}^n
    \phi^{(k)}_i \, \phi^{(k)}_j
    =
    \frac 2 { n^2 }
    \sum_{k = 1}^n
    \Phi^{(k)}_i \, \Phi^{(k)}_j
    -
    \frac 1 { n^2 }
    =
    \frac 1 n \,
    \delta_{i, j}.
  $$
}
%

@@

For the single-bin updating scheme,
we chose $\alpha_0 = 0.0001$
and ran the simulation for $t = 10^8$ steps.
%
As shown in Fig. \ref{fig:errsbin},
the numerical results of the terminal error agreed well with
the analytical prediction by Eq. \eqref{eq:error_invt}.
%
For both global and local sampling processes,
the optimal $\lambda$ occurred around $1.0$,
verifying Eq. \eqref{eq:alpha_invt1}.
%
%Recall that $1/\lambda$ in Eq. \eqref{eq:alpha_invtlambda}
%represents the relative updating magnitude.
%%
%Figure \ref{fig:errsbin} also shows
%an asymmetry of the error curve.
%A smaller than optimal value of
%the updating magnitude, $1/\lambda$,
%tends to increase the error more rapidly
%than a larger than optimal value,
%suggesting that the over-updating
%is generally preferable to the under-updating
%when the optimal value is unknown.
%
The two error functions
differed only by a multiplicative constant,
again agreeing with the theory.
%
With identical eigenvalues,
$\lambda_1 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling process only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the above multiplicative constant.

>>>>

By assuming Eq. \eqref{eq:w_detailedbalance},
we can diagonalize $\mathbf w$ with a set of
eigenvectors
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 1}^n
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, p_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { p_i }
  &=
  \delta_{kl}
  .
\label{eq:eig_orthonormal_rows}
\end{align}

@@

In analogous to Eq. \eqref{eq:x2t_average},
we can write,
from Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:y2_sum},
the error after a period $T$,
%
\begin{equation}
  \Err
  =
  \Err_R + \Err_A,
  \label{eq:error_split}
\end{equation}
%
as the sum of the residual error
(from the decay of the initial error)
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
\begin{equation}
  \Err_A
  =
  \int_0^T \int_0^T
  \sum_{k = 2}^n
    \dot u_k\bigl( q(t) \bigr) \,
    \dot u_k\bigl( q(t') \bigr) \,
    \kappa_k(t - t') \, dt \, dt',
  \label{eq:error_asym}
\end{equation}
%
where
\begin{align*}
  u_k(q')
  &\equiv
  e^{\lambda_k \, [q' - q(T)]},
  \\
  \kappa_k(t - t')
  &\equiv
  \left\langle
    \theta_k(t) \, \theta_k(t')
  \right\rangle.
\end{align*}

@@

In the asymptotic regime
(i.e., if the simulation length, $T$, is sufficiently long),
we can approximate $\kappa_k(t)$ as
%
\begin{equation}
  \kappa_k(t) \approx \Gamma_k \, \delta(t)
  ,
\label{eq:kappa_delta}
\end{equation}
%
where
\begin{equation}
  \Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_k = \sum_{t = -\infty}^\infty \kappa_k(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_asym} is simplified as
%
\begin{align}
  \Err_A
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt.
  \label{eq:error_asym1}
\end{align}
%
In this way,
the underlying sampling process
affects the error only through the set of numbers, $\Gamma_i$'s.

@@

We shall use this result to model the initial error
before starting the schedule $\alpha(t)$ to be optimized.
%
That is, we assume
$\alpha(t) = \alpha_0$
for $t < 0$, and
the system has entered the equilibrium
under a fixed $\alpha_0$ at $t = 0$.
%
We can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.
%
Then,
the residual error at the end of period $T$
is given by
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
      \frac 1 2 \, \alpha_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
Note that the residual error
depends on $\alpha(t)$ only through
the integral $q(T) = \int_0^T \alpha(t) \, dt$.
%
Thus, if we variate $\alpha(t)$ under a fixed $q(T)$,
this residual part can be ignored,
and the optimization can be done exclusively
on the asymptotic part, $\Err_A$.

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
%\label{eq:error_asym_Lagrangian}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  ,
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in th above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

@@

Parallel to Eq. \eqref{eq:phi_pbc},
we define the out-of-boundary values
of $\phi_i$ as
%
\begin{equation}
  \phi_i
  =
  \begin{dcases}
    \phi_{ 1 - i }           & \mathrm{for \;} i \le 0, \\
    \phi_{ 2 \, n + 1 - i }  & \mathrm{for \;} i > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that Eq. \eqref{eq:wmul_to_convol}
still holds.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}%
Thus, we also have $n$ orthonormal eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
compatible with Eq. \eqref{eq:phi_refl},
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, \pi}
            {                    n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
with eigenvalues
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \pi}{n}
  .
\label{eq:wband_eigenvalue_refl}
\end{align}
%
\note{Derivation.
  First consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies Eq. \eqref{eq:phi_refl}.
  %
  So, by Eq. \eqref{eq:wmul_to_convol},
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue_refl}.
}%

@@

For the single-bin updating scheme,
we chose $\alpha_0 = 0.0001$
and ran the simulation for $T = 10^8$ steps.
%
As shown in Fig. \ref{fig:errsbin},
the numerical results of the terminal error agreed well with
the analytical prediction by Eq. \eqref{eq:error_invt}.
%
For both global and local sampling processes,
the optimal $\lambda$ occurred around $1.0$,
verifying Eq. \eqref{eq:alpha_invt1}.
%
%Recall that $1/\lambda$ in Eq. \eqref{eq:alpha_invtlambda}
%represents the relative updating magnitude.
%%
%Figure \ref{fig:errsbin} also shows
%an asymmetry of the error curve.
%A smaller than optimal value of
%the updating magnitude, $1/\lambda$,
%tends to increase the error more rapidly
%than a larger than optimal value,
%suggesting that the over-updating
%is generally preferable to the under-updating
%when the optimal value is unknown.
%
The two error functions
differed only by a multiplicative constant,
again agreeing with the theory.
%
With identical eigenvalues,
$\lambda_1 = \cdots = \lambda_n = 1$,
the error in Eq. \eqref{eq:error_invt}
depends on the sampling process only through
the sum $\sum_{ k = 2 }^n \Gamma_k$,
which gives rise to the above multiplicative constant.
